---
title: "Defining a new distribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining a new distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
```

```{r setup}
library(distributions7)
```

## The idea in one sentence

A new distribution in **distributions7** requires only its **density**
(`distrib_pdf`). Everything else --- cumulative distribution function,
quantile function, random number generator, score (gradient), observed and
expected Hessian, and the moments --- is then available automatically through
numerical fallbacks. A closed-form expression derived later is registered as
a method and transparently takes over, with no change to calling code.

## How dispatch makes this work

Every operation in the package is an S7 *generic* dispatching on the
distribution object:

| Generic | Meaning |
|---|---|
| `distrib_pdf` | density / mass function |
| `distrib_cdf` | cumulative distribution function |
| `distrib_quantile` | quantile function (inverse CDF) |
| `distrib_rng` | random number generator |
| `distrib_gradient` | score: first derivatives of the log-density w.r.t. the parameters |
| `distrib_hessian` | observed Hessian of the log-density |
| `distrib_expected_hessian` | expected Hessian (negative Fisher information) |
| `distrib_deriv3`, `distrib_deriv4` | third- and fourth-order derivatives (observed or expected) |
| `distrib_grad_y`, `distrib_hess_y` | derivatives of the log-density w.r.t. the response `y` (continuous) |
| `mean`, `variance`, `std_dev`, `skewness`, `kurtosis`, `moment` | moments |

Default methods for all of these (except `distrib_pdf`) are registered on the
**base classes** `continuous_distrib` and `discrete_distrib`. Because S7
always selects the most specific applicable method, a subclass inherits those
defaults until something more specific is registered. The defaults are:

* `distrib_cdf` --- numerical integration (continuous) or summation of the pmf
  (discrete) of the density;
* `distrib_quantile` --- root-finding / table inversion on the CDF;
* `distrib_rng` --- Generalized Ratio-of-Uniforms (`rng_grou()`), which needs
  nothing but the density; inverse-transform sampling,
  `distrib_quantile(runif(n))`, is used instead whenever an analytical
  quantile function is supplied;
* `distrib_gradient`, `distrib_hessian` --- central finite differences of the
  log-density (see `numerical_gradient()` / `numerical_hessian()`);
* `distrib_deriv3`, `distrib_deriv4` --- finite differences of the Hessian (see
  `numerical_deriv3()` / `numerical_deriv4()`), with the expected versions
  obtained by `expectation()` of the observed ones;
* `distrib_expected_hessian` --- the `expectation()` of the observed Hessian;
* the moments --- numerical `expectation()` of the relevant power.

## The minimal recipe

A new distribution needs exactly three things:

1. **A class** that inherits from `continuous_distrib` or `discrete_distrib`.
2. **A `distrib_pdf` method** for that class.
3. **A constructor** that fills in the metadata (name, support bounds,
   parameter names, domains and link functions).

Nothing else is required.

### Worked example: the Laplace distribution

The Laplace (double-exponential) density is
$f(y;\mu,b)=\frac{1}{2b}\exp\!\left(-\frac{|y-\mu|}{b}\right)$, with
$\mu\in\mathbb{R}$ and $b>0$. Base R has no `dlaplace`, so this is a genuinely
new distribution.

**Step 1 --- the class.** It is continuous, so we inherit from
`continuous_distrib`:

```{r}
Laplace <- S7::new_class("Laplace", parent = continuous_distrib)
```

**Step 2 --- the density.** A method for `distrib_pdf`. The signature is always
`(distrib, y, theta, log = FALSE)`, where `theta` is a named list of parameters
(each element a scalar or a vector aligned with `y`):

```{r}
S7::method(distrib_pdf, Laplace) <- function(distrib, y, theta, log = FALSE) {
  mu <- theta[[1]]
  b  <- theta[[2]]
  log_d <- -log(2 * b) - abs(y - mu) / b
  if (log) log_d else exp(log_d)
}
```

Two conventions matter. `theta` is indexed positionally (`theta[[1]]`,
`theta[[2]]`) so that the parameter *names* stay free, and the `log` branch is
always implemented, because the numerical machinery evaluates the density on
the log scale for stability.

**Step 3 --- the constructor.** A plain function returning an object of the
class with the metadata filled in:

```{r}
laplace_distrib <- function(link_mu = linkfunctions7::identity_link(),
                            link_b  = linkfunctions7::log_link()) {
  Laplace(
    distrib_name = "laplace",
    dimension    = "univariate",
    bounds       = c(-Inf, Inf),                 # support of Y
    params       = c("mu", "b"),
    params_interpretation = c(mu = "location", b = "scale"),
    n_params     = 2,
    params_bounds = list(mu = c(-Inf, Inf), b = c(0, Inf)),
    link_params   = list(mu = link_mu, b = link_b)
  )
}
```

The `params_bounds` are used by the fallbacks (finite-difference steps are
shrunk so they never step outside a parameter's domain) and by
`generate_random_theta()`. The `link_params` are used by downstream modelling
code; `identity_link()` and `log_link()` come from **linkfunctions7**.

That is the entire definition. Everything now works:

```{r}
d  <- laplace_distrib()
th <- list(mu = 1, b = 2)

# density (what we defined)
distrib_pdf(d, c(0, 1, 3), th)

# CDF, quantile, RNG --- all numerical fallbacks
distrib_cdf(d, c(-1, 1, 4), th)
distrib_quantile(d, c(0.25, 0.5, 0.75), th)   # median is exactly mu = 1
head(distrib_rng(d, 5, th))

# score and Hessian --- finite differences of the log-density
str(distrib_gradient(d, c(-1, 3), th))
str(distrib_hessian(d, c(-1, 3), th))

# moments --- numerical integration against the density
c(mean = mean(d, th), variance = variance(d, th))  # Laplace variance is 2 b^2 = 8
```

We can check the fallbacks against the known closed forms for the Laplace:

```{r}
# analytical Laplace CDF
lap_cdf <- function(q, mu, b) {
  ifelse(q < mu, 0.5 * exp((q - mu) / b), 1 - 0.5 * exp(-(q - mu) / b))
}
q <- c(-3, -1, 0, 1, 4)
max(abs(distrib_cdf(d, q, th) - lap_cdf(q, 1, 2)))   # ~1e-10
```

```{r}
# variance should be exactly 2 b^2
all.equal(variance(d, th), 2 * th$b^2)
```

The object also prints and plots like any built-in distribution:

```{r}
d
```

```{r, fig.width = 6, fig.height = 4}
plot(d, th)
```

## Discrete distributions

For a discrete distribution, inherit from `discrete_distrib` instead and let
`distrib_pdf` be the **probability mass function**. The one extra requirement is
that the support has a **finite lower bound** (`bounds[1]`), which every standard
count distribution satisfies. The discrete fallbacks build a cumulative-mass
table from that lower bound, growing it geometrically as needed.

```{r}
# A shifted-geometric example: P(Y = k) = (1-p) p^k for k = 0, 1, 2, ...
Geom <- S7::new_class("Geom", parent = discrete_distrib)
S7::method(distrib_pdf, Geom) <- function(distrib, y, theta, log = FALSE) {
  p <- theta[[1]]
  log_d <- log(1 - p) + y * log(p)
  if (log) log_d else exp(log_d)
}
geom_distrib <- function(link_p = linkfunctions7::logit_link()) {
  Geom(
    distrib_name = "geometric", dimension = "univariate", bounds = c(0, Inf),
    params = "p", params_interpretation = c(p = "success prob."),
    n_params = 1, params_bounds = list(p = c(0, 1)),
    link_params = list(p = link_p)
  )
}

g <- geom_distrib()
distrib_cdf(g, 0:3, list(p = 0.6))                      # 1 - p^(k+1)
distrib_quantile(g, c(0.5, 0.9), list(p = 0.6))
c(mean = mean(g, list(p = 0.6)))                        # p / (1 - p) = 1.5
```

## Adding closed forms for performance

The numerical fallbacks are convenient but slower and less precise than
analytical formulas. Whenever a closed form is available, registering it as a
method makes it take over automatically, and existing code keeps working.
There is no need to override everything: any subset of the methods can be
supplied.

For the Laplace, all of these are available in closed form. For instance the
CDF, quantile and score:

```{r}
S7::method(distrib_cdf, Laplace) <- function(distrib, q, theta,
                                             lower.tail = TRUE, log.p = FALSE) {
  mu <- theta[[1]]; b <- theta[[2]]
  res <- ifelse(q < mu, 0.5 * exp((q - mu) / b), 1 - 0.5 * exp(-(q - mu) / b))
  if (!lower.tail) res <- 1 - res
  if (log.p) log(res) else res
}

S7::method(distrib_quantile, Laplace) <- function(distrib, p, theta,
                                                  lower.tail = TRUE, log.p = FALSE) {
  mu <- theta[[1]]; b <- theta[[2]]
  if (log.p) p <- exp(p)
  if (!lower.tail) p <- 1 - p
  mu - b * sign(p - 0.5) * log(1 - 2 * abs(p - 0.5))
}

# score: d/dmu = sign(y - mu) / b,  d/db = (|y - mu| / b - 1) / b
S7::method(distrib_gradient, Laplace) <- function(distrib, y, theta,
                                                  scale = c("parameter", "link"), ...) {
  mu <- theta[[1]]; b <- theta[[2]]
  list(
    mu = sign(y - mu) / b,
    b  = (abs(y - mu) / b - 1) / b
  )
}
```

Note the signature of the derivative method: it must mirror the generic, which
carries a `scale` argument (and `...`). **A method always returns
parameter-scale derivatives**; the chain rule is never implemented by hand.
The generic applies the link-scale transformation afterwards when the caller
asks for `scale = "link"`. The same applies to `distrib_hessian`,
`distrib_expected_hessian`, `distrib_deriv3` and `distrib_deriv4`; the
probability-function methods (`distrib_pdf`, `distrib_cdf`, `distrib_quantile`,
`distrib_rng`) keep their plain signatures.

The results are identical to what the fallbacks produced, only faster and exact:

```{r}
max(abs(distrib_cdf(d, q, th) - lap_cdf(q, 1, 2)))       # now exact
str(distrib_gradient(d, c(-1, 3), th))                   # now analytical
```

A practical suggestion for the order in which to add closed forms:

1. **`distrib_gradient`** --- the single biggest speed-up for model fitting,
   since the score is evaluated at every optimisation step. Writing it in C++
   (via **Rcpp**), as the built-in distributions do, helps further.
2. **`distrib_hessian`** --- for Newton-type optimisation and standard errors.
3. **`distrib_cdf` / `distrib_quantile` / `distrib_rng`** --- for fast or
   high-volume simulation and tail probabilities.
4. **`distrib_expected_hessian`** and **analytical moments** --- when a closed
   form exists and the fitting method is Fisher scoring, or exact moments are
   needed.

## Checking analytical derivatives

Hand-written analytical derivatives should be verified against the numerical
reference implementations, which are exported for exactly this purpose:

```{r}
y <- distrib_rng(d, 50, th)

ana <- distrib_gradient(d, y, th)
num <- numerical_gradient(d, y, th)
max(abs(ana$mu - num$mu))
max(abs(ana$b  - num$b))
```

`numerical_gradient()` and `numerical_hessian()` take the same
`(distrib, y, theta)` arguments as the generics and return the finite-difference
estimates, so a component-by-component comparison is a one-liner.

## Validating the distribution

Once a distribution is defined, `check_distrib()` runs a battery of numerical
self-consistency checks: it verifies that the density integrates (or sums) to
one, that the CDF, quantile function and RNG agree with each other, and that
every analytical derivative matches finite differences.

```{r}
check_distrib(laplace_distrib(), theta = list(mu = 1, b = 2),
              n = 40, nsim = 2e4, orders = 1:2)
```

This is the fastest way to catch a mistake in a hand-derived score or Hessian:
an analytical `distrib_gradient` that disagrees with the density makes the
corresponding row come back `FAIL`.

## Derivatives on the link scale

Every parameter carries a link function, and the derivative generics accept
`scale = "link"` to differentiate with respect to the unconstrained linear
predictor $\eta = g(\theta)$ instead of $\theta$ itself:

```{r}
d <- gaussian_distrib()          # identity link on mu, log link on sigma
th <- list(mu = 1.5, sigma = 2)
y <- c(0.4, 1.9, 3.2)

distrib_gradient(d, y, th)                    # d l / d theta
distrib_gradient(d, y, th, scale = "link")    # d l / d eta
```

Because $\sigma = e^{\eta}$, the second element of the link-scale score is simply
the parameter-scale score times $\sigma$. Higher orders follow the Faà di Bruno
chain rule and are available up to order 4; see `?link_scale_derivatives` for the
formulas.

## Fitting a distribution to data

`fit_distrib()` maximises the likelihood on the link scale — where the parameters
are unconstrained — using the analytical score and information, and then reports
the estimates back on the parameter scale:

```{r}
set.seed(1)
d <- gaussian_distrib()
y <- distrib_rng(d, 500, list(mu = 2, sigma = 3))
fit_distrib(d, y)
```

Confidence intervals are built symmetrically on the link scale and mapped through
$g^{-1}$, so they always respect the parameter's domain. For a probability, for
instance, the interval can never leave $(0, 1)$:

```{r}
set.seed(2)
fit_distrib(bernoulli_distrib(), rbinom(50, 1, 0.9))
```

## Non-differentiable parameters

Some distributions have a log-likelihood that is not differentiable in one of its
parameters. The archetype is the **Laplace** distribution
$f(y;\mu,b)=\frac{1}{2b}e^{-|y-\mu|/b}$, whose density has a kink at $y=\mu$: the
score $\partial\ell/\partial\mu=\operatorname{sign}(y-\mu)/b$ exists only almost
everywhere, and the observed curvature $\partial^2\ell/\partial\mu^2$ is **zero**,
so Newton-Raphson cannot update $\mu$.

The package handles this in two ways. First, a distribution can declare which
parameters are smooth via the `params_smooth` metadata (shown in `print()`):

```{r}
d <- laplace_distrib()
param_smoothness(d)
```

Second, and most importantly, `distrib_expected_hessian` returns the **Fisher
information** computed from the variance of the score,
$I=\mathbb{E}[\nabla\ell\,\nabla\ell^\top]$, rather than from $-\mathbb{E}[H]$.
For a regular model the two coincide, by the second Bartlett identity, but
only the score-variance form stays valid when the identity fails. It gives
the correct information $1/b^2$ for the Laplace location, where
$-\mathbb{E}[H]$ would give a degenerate zero:

```{r}
th <- list(mu = 1, b = 2)
distrib_hessian(d, c(-1, 0, 2), th)$mu_mu          # observed curvature: 0
distrib_expected_hessian(d, 0, th)$mu_mu           # Fisher information: -1/b^2
```

This means that even a user-defined non-smooth distribution supplying only its
density (and, ideally, an analytic score at the kink) obtains the right expected
information automatically, because the default `distrib_expected_hessian` uses the
outer product of the score. See `?laplace_distrib` for the full worked example.

## Summary

* **Minimum to define a distribution:** a subclass of `continuous_distrib` or
  `discrete_distrib`, a `distrib_pdf` method, and a constructor with the
  metadata. (Discrete distributions additionally need a finite lower support
  bound.)
* The density alone yields the CDF, quantile function, RNG, score, observed
  and expected Hessian, and all moments, via numerical fallbacks.
* Register analytical methods incrementally for speed and precision; they
  override the fallbacks automatically through S7 dispatch.
* Validate analytical derivatives with `numerical_gradient()` /
  `numerical_hessian()`.
