---
title: "Derivatives and the link scale"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Derivatives and the link scale}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(1)
```

```{r setup}
library(distributions7)
```

Most of what **distributions7** offers beyond densities is derivatives: the
score, the information, and the third and fourth derivatives of the
log-likelihood. This vignette explains what they are, the two scales they are
available on, and why the second scale --- the link scale --- is the one an
optimizer needs.

## The derivatives of the log-likelihood

For one observation, write $\ell(\theta) = \log f(y; \theta)$. The package
returns its derivatives with respect to the parameters, per observation, as
named lists:

```{r}
d <- gaussian1_distrib()
theta <- list(mu = 2, sigma = 3)
y <- distrib_rng(d, 4, theta)

distrib_gradient(d, y, theta)          # score: d ell / d theta
```

```{r}
distrib_hessian(d, y, theta)           # observed: d^2 ell / d theta^2
```

The **expected** Hessian --- the negative Fisher information --- is a separate
generic, because for a correct model it does not depend on $y$:

```{r}
distrib_expected_hessian(d, 0, theta)
```

Third and fourth derivatives are there too, keyed by their multi-index, for
anything that needs curvature beyond the quadratic:

```{r}
distrib_deriv3(d, y, theta)
```

These are exact, computed by closed-form C++ kernels, not by differencing. A
separate pair, `distrib_grad_y()` and `distrib_hess_y()`, differentiates with
respect to the response $y$ instead of the parameters.

## Two scales

Every derivative generic takes a `scale` argument. The default,
`scale = "parameter"`, is what the examples above used: derivatives with
respect to $\theta$ on its natural, constrained scale.

`scale = "link"` instead gives derivatives with respect to the **unconstrained**
parameter $\eta = g(\theta)$ behind each link. For the Gaussian the default link
on $\sigma$ is the log, so $\eta_\sigma = \log\sigma$:

```{r}
distrib_gradient(d, y, theta, scale = "link")
```

Nothing was differentiated numerically to get this. The chain rule was applied
to the parameter-scale derivatives, exactly. For a single parameter with link
$g$, $\partial\ell/\partial\eta = (\partial\ell/\partial\theta)\,/\,g'(\theta)$;
with the log link on $\sigma$ that factor is simply $\sigma$:

```{r}
g_par  <- distrib_gradient(d, y, theta)
g_link <- distrib_gradient(d, y, theta, scale = "link")

all.equal(g_par$sigma * theta$sigma, g_link$sigma)
```

Under an identity link the two scales coincide, as they must:

```{r}
d_id <- gaussian1_distrib(
  link_mu    = linkfunctions7::identity_link(),
  link_sigma = linkfunctions7::identity_link()
)
all.equal(
  distrib_gradient(d_id, y, theta),
  distrib_gradient(d_id, y, theta, scale = "link")
)
```

The transformation works at every order. The general rule is Faà di Bruno's
formula --- the chain rule for higher derivatives --- applied with a diagonal
Jacobian, since each parameter has its own link. It is written once, in the body
of each generic, so `scale = "link"` is available uniformly:

```{r}
distrib_deriv3(d, y, theta, scale = "link")
```

## Why the link scale is the useful one

A standard deviation is positive, a probability lives in $(0, 1)$, a shape
parameter is positive. Maximizing a likelihood over such constrained parameters
means either fighting the boundary or bolting on constraints. Reparameterise
through the link and the problem becomes unconstrained: $\eta$ ranges over the
whole real line, and a Newton or Fisher-scoring step needs exactly the
link-scale score and information this package computes in closed form.

That is precisely what `fit_distrib()` does --- it optimizes on the link scale
and maps back --- and it is why its confidence intervals cannot leave a
parameter's domain. See `vignette("fitting-a-model")` for that side of the story.

The information transforms consistently. Write $h = g^{-1}$, so that
$h'(\eta) = 1/g'(\theta)$ is the derivative of the inverse link — the same
factor that already appeared in the first-order rule above. Because the score
has zero expectation, the expected information on the link scale is the
sandwich $\operatorname{diag}(h') \, \mathcal{I} \, \operatorname{diag}(h')$;
for the log link on $\sigma$ we have $h'(\eta) = e^{\eta} = \sigma$, so the
$\sigma\sigma$ entry is multiplied by $\sigma^2$:

```{r}
eh_par  <- distrib_expected_hessian(d, 0, theta)
eh_link <- distrib_expected_hessian(d, 0, theta, scale = "link")

c(parameter = eh_par$sigma_sigma[1],
  scaled    = theta$sigma^2 * eh_par$sigma_sigma[1],
  link      = eh_link$sigma_sigma[1])
```

## Expected higher derivatives, when there is no formula

`distrib_deriv3()` and `distrib_deriv4()` also take `expected = TRUE`. Some
distributions have closed-form expected derivatives; where they do not, the
value is approximated, and `approx` chooses how:

```{r}
distrib_deriv3(gamma2_distrib(), 0, list(mu = 3, sigma2 = 2), expected = TRUE)
```

The strategies are `"bartlett"`, `"integrate"` and `"mc"`, trading speed against
robustness. They rest on one idea: the outer product of gradients is the
order-two case of the Bartlett identities, and the general identity expresses an
expected derivative of any order through products of lower-order ones. `?distrib_deriv3`
has the details; the default is chosen to be both cheap and valid, including for
models where the usual regularity assumptions fail.

## A note for authors of derivative methods

A user-written analytical derivative method **must** include the `scale`
argument in its signature, even if the method only ever returns the parameter
scale:

```{r eval = FALSE}
S7::method(distrib_gradient, MyDist) <- function(distrib, y, theta,
                                                 scale = c("parameter", "link"), ...) {
  # ... return the parameter-scale gradient; the generic handles "link" itself
}
```

S7 requires a method's formals to include the generic's named arguments, and the
link-scale transformation lives in the generic, wrapping whatever the method
returns. Omit `scale` and dispatch will fail. The numerical fallbacks and all
built-in methods already follow this pattern.

## Where to go next

- `vignette("fitting-a-model")` --- the link scale put to work in estimation.
- `vignette("defining-a-distribution")` --- adding a distribution that inherits
  all of this.
- `?distrib_gradient`, `?distrib_deriv3` --- the full reference.
