---
title: "Fitting a distribution to data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting a distribution to data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
set.seed(1)
```

```{r setup}
library(distributions7)
```

This vignette walks through fitting a distribution to a sample by maximum
likelihood, reading the result, checking it against the data, and simulating
from it. It assumes familiarity with distribution objects;
`vignette("defining-a-distribution")` is the place to start otherwise.

## A first fit

Every distribution has a constructor. It takes the link functions for its
parameters, but the defaults are sensible, so most of the time it is called
with no arguments:

```{r}
d <- gamma_distrib()
d
```

We simulate some data from a known parameter value and fit the distribution back
to it. Parameters travel as a named list.

```{r}
y <- distrib_rng(d, 500, list(mu = 3, sigma2 = 2))
fit <- fit_distrib(d, y)
fit
```

The estimates recover the values the sample was generated from, and the print
method shows them twice: on the **parameter scale**, and on the **link
scale** where the fitting actually happened. More on that split below.

## Reading the fit

`fit_distrib()` returns an object that answers the familiar extractors:

```{r}
coef(fit)
vcov(fit)
logLik(fit)
```

`coef()` and `vcov()` also take a `scale` argument, which returns the same
quantities on the unconstrained scale:

```{r}
coef(fit, scale = "link")
```

The object carries confidence intervals, the information criteria, the number of
iterations, and the method that was actually used --- all visible in the print
above.

## Why the link scale

A gamma's mean and variance are positive; a beta's probability lives in
$(0, 1)$; a Gaussian's standard deviation is positive. Optimising those
constrained parameters directly means fighting the boundary. Instead
`fit_distrib()` optimises the **unconstrained** parameters behind each link ---
$\log\mu$ and $\log\sigma^2$ for the gamma --- where there is no boundary to hit,
using the exact score and information the package provides on that scale.

The estimates are then mapped back, and so are the intervals. Because a
confidence interval is built symmetrically on the link scale and pushed through
$g^{-1}$, its endpoints can never leave the parameter's domain:

```{r}
b <- bernoulli_distrib()
fit_distrib(b, rbinom(40, 1, 0.9))
```

Even with a probability estimate close to one and only forty observations, the
interval stays inside $(0, 1)$ --- a symmetric interval on the probability scale
would have spilled past it.

## Choosing how it is optimised

The default is Fisher scoring, which uses the *expected* information. Two other
methods are available, and Fisher scoring and Newton fall back to BFGS if they
fail to converge:

```{r}
fit_newton <- fit_distrib(d, y, method = "newton")
fit_bfgs   <- fit_distrib(d, y, method = "bfgs")

c(scoring = as.numeric(logLik(fit)),
  newton  = as.numeric(logLik(fit_newton)),
  bfgs    = as.numeric(logLik(fit_bfgs)))
```

They agree on the maximum. Fisher scoring is the default because it uses the
expected information, which stays well-behaved even where the observed Hessian
does not, and that matters for distributions with a non-smooth parameter such
as the Laplace.

The optimisation runs through
[optimizers7](https://statmodels7.github.io/optimizers7/), and `method` also
accepts an optimiser object from that package, which is then used as given.
Choosing the optimiser this way also chooses its stopping rule, its line
search and its iteration budget:

```{r}
library(optimizers7)

fit_lbfgs <- fit_distrib(d, y, method = lbfgs(criterion = crit_grad(1e-12)))
fit_nm    <- fit_distrib(d, y, method = nelder_mead(maxit = 5000))

c(lbfgs = as.numeric(logLik(fit_lbfgs)),
  nelder_mead = as.numeric(logLik(fit_nm)))
c(fit_lbfgs@method, fit_nm@method)
```

The three named strategies fall back to BFGS when they fail to converge; an
optimiser supplied explicitly is never replaced, so the method reported is
always the method that produced the estimates.

## Checking the fit against the data

A fit remembers the data it came from, so it can be drawn over it. For a
continuous distribution `plot()` shows a kernel density of the sample with the
fitted density on top:

```{r}
plot(fit)
```

For a discrete distribution it shows the observed frequencies as bars with the
fitted probability mass overlaid, because a kernel density would misrepresent
counts:

```{r}
p <- poisson_distrib()
yp <- distrib_rng(p, 300, list(mu = 4))
plot(fit_distrib(p, yp))
```

## Simulating from the fit

`simulate()` draws new samples from the fitted distribution. Each replicate has
the same length as the original data, which makes it a one-liner to build a
parametric bootstrap of any statistic:

```{r}
sims <- simulate(fit, 200, seed = 1)
dim(sims)

boot_median <- vapply(sims, median, numeric(1))
quantile(boot_median, c(0.025, 0.975))
```

It follows the `stats::simulate()` contract: a supplied `seed` makes the draw
reproducible and is reported back on the result, and the caller's random stream
is left untouched.

## Validating a distribution numerically

Before a fit is trusted, and especially before a user-written distribution
is, `check_distrib()` puts it through thirteen numerical checks.
It confirms that the density integrates to one, that the distribution function
agrees with the density, that the quantile inverts it, that the generator
follows it, and that every analytical derivative matches finite differences, on
both the parameter and the link scale:

```{r}
invisible(check_distrib(d, list(mu = 3, sigma2 = 2)))
```

Had a mistake crept into an analytical derivative, the corresponding line would
read `FAIL`, with the size of the discrepancy printed next to it. It is the
fastest way to catch a wrong formula.

## Further reading

- `vignette("defining-a-distribution")` --- defining a new distribution, which
  then works with everything shown here.
- `?fit_distrib`, `?check_distrib` --- the full argument lists.
- `?distrib_gradient` --- the derivatives, including what `scale = "link"` does.
