---
title: "Fitting a distribution to data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting a distribution to data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
set.seed(1)
```

```{r setup}
library(distributions7)
```

This vignette walks through fitting a distribution to a sample by maximum
likelihood, reading the result, checking it against the data, and simulating
from it. It assumes you already know what a distribution object is; if not,
`vignette("defining-a-distribution")` is the place to start.

## A first fit

Every distribution has a constructor. It takes the link functions for its
parameters, but the defaults are sensible, so most of the time you call it with
no arguments:

```{r}
d <- gamma_distrib()
d
```

We simulate some data from a known parameter value and fit the distribution back
to it. Parameters travel as a named list.

```{r}
y <- distrib_rng(d, 500, list(mu = 3, sigma2 = 2))
fit <- fit_distrib(d, y)
fit
```

The estimates recover the values we generated from, and the print method shows
them twice: on the **parameter scale** you asked about, and on the **link scale**
where the fitting actually happened. More on that split below.

## Reading the fit

`fit_distrib()` returns an object you interrogate with the familiar extractors:

```{r}
coef(fit)
vcov(fit)
logLik(fit)
```

`coef()` and `vcov()` also take a `scale` argument, in case you want the
quantities on the unconstrained scale:

```{r}
coef(fit, scale = "link")
```

The object carries confidence intervals, the information criteria, the number of
iterations, and the method that was actually used --- all visible in the print
above.

## Why the link scale

A gamma's mean and variance are positive; a beta's probability lives in
$(0, 1)$; a Gaussian's standard deviation is positive. Optimising those
constrained parameters directly means fighting the boundary. Instead
`fit_distrib()` optimises the **unconstrained** parameters behind each link ---
$\log\mu$ and $\log\sigma^2$ for the gamma --- where there is no boundary to hit,
using the exact score and information the package provides on that scale.

The estimates are then mapped back, and so are the intervals. Because a
confidence interval is built symmetrically on the link scale and pushed through
$g^{-1}$, its endpoints can never leave the parameter's domain:

```{r}
b <- bernoulli_distrib()
fit_distrib(b, rbinom(40, 1, 0.9))
```

Even with a probability estimate close to one and only forty observations, the
interval stays inside $(0, 1)$ --- a symmetric interval on the probability scale
would have spilled past it.

## Choosing how it is optimised

The default is Fisher scoring, which uses the *expected* information. Two other
methods are available, and Fisher scoring and Newton fall back to BFGS if they
fail to converge:

```{r}
fit_newton <- fit_distrib(d, y, method = "newton")
fit_bfgs   <- fit_distrib(d, y, method = "bfgs")

c(scoring = as.numeric(logLik(fit)),
  newton  = as.numeric(logLik(fit_newton)),
  bfgs    = as.numeric(logLik(fit_bfgs)))
```

They agree on the maximum. Fisher scoring is the default because it uses the
expected information, which stays well-behaved even where the observed Hessian
does not --- a point that matters for distributions with a non-smooth
parameter, such as the Laplace.

## Checking the fit against the data

A fit remembers the data it came from, so it can be drawn over it. For a
continuous distribution `plot()` shows a kernel density of the sample with the
fitted density on top:

```{r}
plot(fit)
```

For a discrete distribution it shows the observed frequencies as bars with the
fitted probability mass overlaid, because a kernel density would misrepresent
counts:

```{r}
p <- poisson_distrib()
yp <- distrib_rng(p, 300, list(mu = 4))
plot(fit_distrib(p, yp))
```

## Simulating from the fit

`simulate()` draws new samples from the fitted distribution. Each replicate has
the same length as the original data, which makes it a one-liner to build a
parametric bootstrap of any statistic:

```{r}
sims <- simulate(fit, 200, seed = 1)
dim(sims)

boot_median <- vapply(sims, median, numeric(1))
quantile(boot_median, c(0.025, 0.975))
```

It follows the `stats::simulate()` contract: a supplied `seed` makes the draw
reproducible and is reported back on the result, and the caller's random stream
is left untouched.

## Validating a distribution numerically

Before trusting a fit --- and especially before trusting a distribution you
wrote yourself --- `check_distrib()` puts it through thirteen numerical checks.
It confirms that the density integrates to one, that the distribution function
agrees with the density, that the quantile inverts it, that the generator
follows it, and that every analytical derivative matches finite differences, on
both the parameter and the link scale:

```{r}
invisible(check_distrib(d, list(mu = 3, sigma2 = 2)))
```

If you had made a mistake in an analytical derivative, the corresponding line
would turn red, with the size of the discrepancy next to it. It is the fastest
way to catch a wrong formula.

## Where to go next

- `vignette("defining-a-distribution")` --- add your own distribution, which
  then works with everything shown here.
- `?fit_distrib`, `?check_distrib` --- the full argument lists.
- `?distrib_gradient` --- the derivatives, including what `scale = "link"` does.
